---
title: "Using EvolutionaryGames"
author:
- name: Daniel Gebele
  affiliation: 
  - &id Hochschule Kempten
  email: dngebele@gmail.com
- name: Jochen Staudacher
  affiliation: 
  - &id Hochschule Kempten
  email: jochen.staudacher@hs-kempten.de
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using EvolutionaryGames}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
abstract: "This document gives a few use cases for the EvolutionaryGames 
package. EvolutionaryGames provides basic concepts of evolutionary game 
theory, like e.g. finding evolutionary stable strategies and computing 
and drawing evolutionarily stable sets as well as phase diagrams for 
various evolutionary dynamics for single-population games with two, 
three and four different phenotypes.\n\n**Keywords**: 
Evolutionary game theory, evolutionary stable strategies, 
evolutionarily stable sets, evolutionary game dynamics."
bibliography: references.bib
biblio-style: apsr
---


## Motivation 
The field of evolutionary game theory is concerned with interactions 
among large populations of strategically interacting agents. 
These agents may adapt their behaviour according to current 
payoff opportunities. Evolutionary game theory originates 
from biology [@ESS], but nowadays has become an established 
tool in other disciplines, in particular in 
economics (see e.g. [@SandholmBook]) 
and computer science (see e.g. [@SuriCompEvol]).

The purpose of our package EvolutionaryGames is to enhance 
the R ecosystem by tools to plot and present evolutionary game dynamics 
in a both informative and attractive way. One similar tool 
in the public domain is William Sandholm's package dynamo [@dynamo] 
written in Mathematica. Dynamo [@dynamo] frequently served 
as an inspiration for our implementation. Still, even though 
dynamo [@dynamo] itself is free, Mathematica is 
proprietary software and it is nowhere near as widely used 
as R. Also, the package structures in R and the widespread 
availability of packages via CRAN make it much easier for 
the user to employ parts of our package EvolutionaryGames in his or her 
own software.

The following very brief presentation of concepts in evolutionary game 
theory and how EvolutionaryGames addresses them mainly follows 
the two books by Weibull [@WeibullBook] and Sandholm 
[@SandholmBook]. Furthermore, our vignette provides 
detailed references to the respective original articles. 
Our point is to show how various concepts from 
evolutionary game theory can be computed via the package 
EvolutionaryGames and where to find the conceptual, 
technical and mathematical details. 
In other words, we would like to stress that this 
little document is by no means intended to serve as an introduction 
to the fascinating field of evolutionary game theory itself. Again, 
for the latter the reader is referred to the the two books by Weibull 
[@WeibullBook] and Sandholm [@SandholmBook]. For an in-depth mathematical 
discussion we also recommend the book by Hofbauer and 
Sigmund [@HofbauerSigmund].

The package EvolutionaryGames focusses on single-population games 
with two, three or four phenotypes. (The first author hopes 
to address multi-population games in a separate package based 
on EvolutionaryGames in the future.)
For single-population games we regard 
EvolutionaryGames as a powerful and attractive alternative to 
William Sandholm's Mathematica package dynamo [@dynamo]. 

## Computing Evolutionary Stable Strategies (ESS)

The concept of evolutionary stable strategies (ESS) was first 
proposed by Maynard Smith and Price in [@ESS]. An incumbent 
strategy $x$ is said to be evolutionary stable if 

- it is either a unique best reply to itself or   
- in case there exists an alternative best reply $y$ to $x$,
then $x$ is a better reply to this mutant strategy $y$ than 
$y$ is to itself.  

If $x$ is a strict Nash equilibrium, then $x$ must also be an ESS.  
Using our package EvolutionaryGames one can compute ESS 
for both $2$ x $2$ and $3$ x $3$ symmetric matrix games.
The function **ESS.R** receives three input arguments:

- `A` : A matrix specifying the symmetric game
- `strategies`: A vector of strings of length $2$ or $3$ 
specifying the names of all strategies
- `floats`: A logical value that handles number representation. If floats 
is set to `TRUE` (default), a floating-point number will be used for the 
output, otherwise the output will be 
specified as fractions.

Let us first take a look at a classical Hawk-Dove game.

```{r, fig.show='hold'}
library(EvolutionaryGames)
ESS(matrix(c(-1, 4, 0, 2), 2, byrow=TRUE), c("Hawk", "Dove"))
ESS(matrix(c(-1, 4, 0, 2), 2, byrow=TRUE), c("Hawk", "Dove"), FALSE)
```

It is well known that there are games which do not possess 
an ESS. A classical example is the game Rock-Scissors-Paper.
In such a case our function **ESS.R** will return NULL.

```{r, fig.show='hold'}
library(EvolutionaryGames)
(A <- matrix(c(1, 2, 0, 0, 1, 2, 2, 0, 1), 3, byrow=T))
ESS(A, c("Rock", "Scissors", "Paper"))
```


## Computing and drawing evolutionarily stable sets

The concept of evolutionarily stable sets 
was first discussed by B. Thomas in [@ESsets]. 
A very readable introduction can also be found in the book 
by Weibull [@WeibullBook], section 2.4.1.  
An evolutionarily stable set is a nonempty closed set 
of symmetric Nash equilibrium strategies $X$ such that 
each strategy $x \in X$ earns at least the same payoff 
against any nearby alternative best reply $y$ as 
$y$ earns against itself with equal payoffs limited 
to the case $y \in X$. Note that any evolutionary stable 
strategy (ESS) constitutes an evolutionarily stable set and 
that the union of evolutionarily stable sets is again an 
evolutionarily stable set. See the book by Weibull [@WeibullBook], 
section 2.4.1, for further details.  
Evolutionarily stable sets are not easy to compute and to plot.
The package EvolutionaryGames computes evolutionarily stable sets of a 
game with two players and three strategies in the case that the game has 
an evolutionary stable strategy (ESS). If the two player three strategy 
game has no ESS, then the code returns a message stating that 
our algorithm cannot calculate evolutionarily stable sets for models 
that do not have a proper ESS. The authors are very well aware 
that there are games having evolutionarily stable sets but no proper ESS.
Still, as our package is not devoted to finding all symmetric 
Nash equilibria of a game and as there currently is no package 
for this task on CRAN, we decided only to handle the case of 
games with two players and three strategies possessing at least one 
proper ESS. Within our algorithm, we need a proper ESS as a 
starting point for our computations. The authors feel that the 
possibility of computing and drawing evolutionarily stable sets in 
such cases is precious for teaching (and research) purposes. However, 
note that computing and drawing evolutionarily stable sets is 
time consuming and is clearly the most elaborate task currently 
performed by the package EvolutionaryGames. Finally, here is an 
example of an evolutionarily stable set 
together with a plot 
including sample trajectories (-- the example itself is 
taken from the book [@Broom2013], p. 52):

```{r, fig.show='hold'}
library(EvolutionaryGames)
A <- matrix(c(-2, 5, 10/9, 0, 5/2, 10/9, -10/9, 35/9, 10/9), 3, byrow=TRUE)
strategies <- c("Hawk", "Dove", "Mixed ESS")
ESS(A, strategies)
ESset(A, strategies)
```


## The various evolutionary dynamics available in EvolutionaryGames

The basic game-theoretic model of biological natural selection is 
the replicator dynamic [@Replicator]. Still, various alternative 
dynamics have been proposed and investigated for different applications. 
Our package currently focusses on continuous dynamics only. 
The following itemization states which evolutionary dynamics are 
currently available in our package EvolutionaryGames:

- **Replicator.R** : The standard replicator dynamic [@Replicator]
- **BNN.R** : The Brown-von Neumann-Nash dynamic [@BNN]
- **BR.R** : The Best Reponse dynamic by Gilboa and Matsui [@BR]
- **ILogit.R** : The imitative Logit dynamic by Weibull [@WeibullBook]. 
It requires the parameter `eta`. 
- **Logit.R** : The Logit dynamic by Fudenberg and Levine [@Logit]. It 
requires the parameter `eta`.
- **MSReplicator.R** : The Maynard-Smith dynamic [@MSReplicator]
- **Smith.R** : The Smith dynamic [@Smith]

## Drawing phase diagrams for two, three and four strategies

### Two Strategies
Drawing phase diagrams for single-population games 
with two, three or 
four phenotypes with different dynamics is the main feature of 
the package EvolutionaryGames.
We start with a rather simple Hawk-Dove game we studied before
when analyzing for its ESS. Using **phaseDiagram2S.R** we 
obtain a phase diagram for the 
population share of hawks invading a population of doves under 
the replicator dynamics.

```{r, fig.show='hold'}
library(EvolutionaryGames)
A <- matrix(c(-1, 4, 0, 2), 2, byrow=TRUE)
phaseDiagram2S(A, Replicator, strategies = c("Hawk", "Dove"))
```

For similar phase diagrams, see e.g. the book by Peters [@PetersBook].  

### Three Strategies
In the phase diagrams for the three strategies the user 
may specify the following parameters for **phaseDiagram3S.R**:

- `A`: A Numeric matrix of size $3$ x $3$ representing the number of 
strategies of a symmetric matrix game.    
- `dynamic`: A function representing an evolutionary dynamic.  
- `params`: A numeric vector with additional parameters for the 
evolutionary dynamic, like e.g. in the cases of the Logit or ILogit 
dynamics.  
- `trajectories`: A numeric matrix of size $m$ x $3$. Each row represents 
the initial values for the trajectory to be examined.  
- `contour`: A logical value that handles contour diagram presentation. 
By default `FALSE`, a nicely coloured contour plot will only be 
shown if the user sets `contour = TRUE`.  
- `vectorField`: A logical value that handles vector field presentation. 
By default `FALSE`, a vector field will only be 
shown if the user sets `vectorField = TRUE`. 
- `strategies`: A vector of strings of length $3$ 
specifying the names of all strategies. By default 
`strategies = c("1","2","3")`. 

In the following we see three plots of the game Rock-Scissors-Paper 
under the Replicator dynamics with the same initial state:

```{r, fig.show='hold'}
library(EvolutionaryGames)
A <- matrix(c(1, 2, 0, 0, 1, 2, 2, 0, 1), 3, byrow=T)
state <- matrix(c(0.7, 0.2, 0.1), 1, 3, byrow=TRUE)
RSP <- c("Rock", "Scissors", "Paper")
phaseDiagram3S(A, Replicator, NULL, state, FALSE, FALSE, strategies = RSP)
phaseDiagram3S(A, Replicator, NULL, state, TRUE, FALSE, strategies = RSP)
phaseDiagram3S(A, Replicator, NULL, state, TRUE, TRUE, strategies = RSP)
```

### Four strategies
Finally, there is also the function **phaseDiagram4S.R** for the case 
of a symmetric matrix game with four strategies. Its parameters 
are rather similar to those of **phaseDiagram3S.R** 
except for the facts that `trajectory` is a numeric vector 
of size $4$ (rather than a matrix) and that there is 
no possibility to enable contour plots or to plot 
vector fields. Instead, there is an additional logical value 
`noRGL` handling diagram rotation. By default `TRUE`, the diagram 
will only be constructed using rgl and thus 
become rotatable if the user specifies `noRGL = FALSE`. 
Finally, we present an example using the Smith dynamic. 

```{r, fig.show='hold'}
library(EvolutionaryGames)
A <- matrix(c(5, -9, 6, 8, 20, 1, 2, -18, -14, 0, 2, 20, 13, 0, 4, -13), 4, 4, byrow=TRUE)
state <- c(0.6, 0.15, 0.1, 0.15)
phaseDiagram4S(A, Smith, NULL, state, noRGL=TRUE)
```

## How to use our package and write your own dynamics

EvolutionaryGames offers you to create your own dynamics. 
In particular, it is easy to write your own continuous dynamics. 
First of all, a  dynamic is nothing other than a function that is passed 
as a parameter to the corresponding function for creating phase diagrams. 
The following code fragment shows you the minimum necessary structure of 
an arbitrary dynamic:

```{r, fig.show='hold' }
MyDynamic <- function (time, state, parameters) {
  
#...
  
  return(list (dX))
}
```

Any dynamic requires the parameters `time`, `state` and `parameters`.
While the parameter `time` is used internally by deSolve to 
solve the initial value problem, the other two parameters 
`state` and `parameters` are used to specify the desired dynamic and are 
available as numeric vectors. In this context, `state` stands for the 
desired initial state under which the model is to be simulated and 
`parameters` contains further parameters, such as the symmetric matrix 
specifying the game and, depending on the dynamic, noise levels or 
similar parameters.

As can be seen in the example, the return value of a specified dynamic 
has to be a numerical list. Each component represents the corresponding 
rate of change of a phenotype under the respective dynamic. 


We now show how to implement a very well known dynamic, the replicator 
dynamics. Our function definition is as follows: 

```{r, fig.show='hold' }
Replicator <- function (time, state, parameters) {
  
#...
  
  return(list (dX))
}
```

The above game and an initial value is passed as a parameter to a function
for the generation of phase diagrams and made retrievable within the 
dynamic. A small limitation, however, is that our matrix needs 
to be converted into a vector due to certain constraints in our 
internal usage of the package deSolve. We recommend that you first 
transfer this vector back into a matrix and maintain the original game. 
This can be done as follows:


```{r, fig.show='hold' }
Replicator <- function (time, state, parameters) {
  a <- parameters
  states <- sqrt(length(a))
  A <- matrix(a, states, byrow = TRUE)
  A <- t(A) # original symmetric game
  
  return(list (dX))
}
```

We now come to the actual part of the implementation. We first calculate 
the rate of change of each phenotype depending on the other phenotypes. 
Immediately afterwards, we calculate the average fitness of 
each phenotype  and then set up the actual replicator dynamics:

```{r, fig.show='hold' }
Replicator <- function(time, state, parameters) {
  a <- parameters
  states <- sqrt(length(a))
  A <- matrix(a, states, byrow = TRUE)
  A <- t(A)

  dX <- c()

  for(i in 1:states) {
    dX[i] <- sum(state * A[i, ])
  }

  avgFitness <- sum(dX * state)

  for(i in 1:states) {
    dX[i] <- state[i] * (dX[i] - avgFitness)
  }

  return(list(dX))
}
```

Our dynamics is now applicable within the predefined functions for 
generating phase diagrams.

## Packages used in EvolutionaryGames
We made use of various packages of the CRAN ecosystem. 
In particular, it was of paramount importance not to write any 
differential equation solvers ourselves, but to make use of 
an established solver, instead.

### deSolve
The R-package deSolve [@deSolve] has long been established as a powerful 
solver for differential equations. In our context, deSolve is used 
to generate the data needed in order to visualize the time evolution 
of a given game under a certain dynamic, i.e. 
we use it for obtaining the correct input for the visualization of 
trajectories in our phase diagrams.

### rgl
We use the package rgl [@rgl], i.e. the R interface to OpenGL, for the 
four strategies case, because here the phase diagram represents a 
three-dimensional simplex. Our users have the possibility to follow 
the development of a game under a given dynamic for all phenotypes by 
rotating the three-dimensional simplex generated by rgl.

### geometry 
The R package geometry [@geometry] helped us with the conversion from 
barycentric to cartesian coordinates for drawing trajectories. Solving 
the initial value problem returns the rates of change of the model, 
which have to be converted into cartesian coordinates before they can be 
drawn into the phase diagram in order to finally form a trajectory.

## Literature